Alternating Automata Modulo First Order Theories

### 2019-02-07 quantifiers/appendix.tex
ok

\section{Emptiness Semi-algorithm for First Order Alternating Automata}
\label{app:algorithm}

\section{Proofs}
\label{app:proofs}

>>>>

\section{Emptiness Checking for First Order Alternating Automata}
\label{app:algorithm}

The execution of Algorithm \ref{alg:impact} consists of three phases,
corresponding to the \textsc{Close}, \textsc{Refine} and
\textsc{Expand} of the original \impact\ procedure \cite{McMillan06}.

\begin{algorithm}[htb]
\begin{algorithmic}[0] \State {\bf input}: a first order
  alternating automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$

  \State {\bf output}: $\top$ if $L(\A)=\emptyset$, or word $w \in
  L(\A)$, otherwise

  \State {\bf data structures}: $\worklist$ and unfolding tree
  $\mathcal{U} = \tuple{N,E,\rootNode,U,\lhd}$,
  where: \begin{compactitem}
  \item $N$ is a set of nodes,
  \item $E \subseteq N \times \Sigma \times N$ is a set of edges labeled by input events,
  \item $U : N \rightarrow \posforms(Q,\emptyset)$ is a labeling of
    nodes with positive sentences
  \item $\lhd \subseteq N \times N$ is a coverage relation,
  \end{compactitem}
  \State {\bf initially} $\worklist = \set{\rootNode}$ and $N=E=U=\lhd=\emptyset$.
\end{algorithmic}
\begin{algorithmic}[1]   
  \While{$\worklist \neq \emptyset$}
  \label{ln:impact-while}

  \State dequeue $n$ from $\worklist$ 
  \label{ln:impact-dequeue}

  \State $N \leftarrow N \cup \set{n}$

  \State let $\alpha(n) = a_1, \ldots, a_k$ be the input event sequence
  on the path from $\rootNode$ to $n$

  \If{$\substform{\alpha}(\stamp{X}{1}, \ldots, \stamp{X}{k})$ is satisfiable} 
  \Comment{counterexample is feasible}

  \State get model $\nu$ of $\substform{\alpha}(\stamp{X}{1}, \ldots, \stamp{X}{k})$

  \State {\bf return} $w = (a_1,\nu(\stamp{X}{1})) \ldots (a_k,\nu(\stamp{X}{k}))$
  \label{ln:impact-nonempty}
  \Comment{$w \in L(\A)$ by construction}

  \Else \Comment{spurious counterexample}

  \State let $(I_0,\ldots,I_k)$ be a GLI for $\alpha$
  \label{ln:refine-begin}

  \State $b \leftarrow \bot$

  \For{$i=0,\ldots,k$}

  \If{$U(n_i) \not\models I_i$}

  \State $\lhd \leftarrow \lhd \setminus \set{ m \in N \mid (m,n_i) \in \lhd }$
  \Comment{uncover the nodes covered by $n_i$}

  \State $U(n_i) \leftarrow U(n_i) \wedge J_i$
  \Comment{strenghten the label of $n_i$ (Lemma \ref{lemma:refinement})}

  \If{$\neg b$}

  \State $b \leftarrow \Call{Close}{n_i}$
  \label{ln:refine-end}

  \EndIf % $\neg b$

  \EndIf % $\Lambda(n_i) \not\models I_i$

  \EndFor

  \EndIf % $\mu(n)$ is satisfiable

  \If{$n$ is not covered}

  \For{$a \in \Sigma$} 
  \Comment{expand $n$}
  \label{ln:expand-begin}

  \State let $s$ be a fresh node and $e = (n,a,s)$ be a new edge

  \State $E \leftarrow E \cup \set{e}$
  \label{ln:edge-insert}

  \State $U \leftarrow U \cup \set{(s,\top)}$

  \State enqueue $s$ into $\worklist$
  \label{ln:expand-end}

  \EndFor

  \EndIf % $n$ is not covered

  \EndWhile

  \State {\bf return} $\top$
\end{algorithmic}
\caption{\impact-based Semi-algorithm for First Order Alternating Automata}
\label{alg:impact}
\end{algorithm}

\begin{algorithm}[htb]
\begin{algorithmic}[1]
\Function{Close}{$x$} {\bf returns} $\Bool$

\For{$y \in N$ such that $\alpha(y) \prec^* \alpha(x)$}

\If{$U(x) \models U(y)$}

\State $\lhd \leftarrow \left(\lhd \setminus \set{ (p,q) \in \lhd \mid
  \text{$q$ is $x$ or a successor of $x$}}\right) \cup \set{(x,y)}$
\label{ln:close-uncover}

\State {\bf return} $\top$

\EndIf

\EndFor

\State {\bf return} $\bot$

\EndFunction
\end{algorithmic}
\caption{The \textsc{Close} function from Algorithm \ref{alg:impact}}
\end{algorithm}

Let $n$ be a node removed from the worklist at line
\ref{ln:impact-dequeue} and let $\alpha(n)$ be the input sequence
labeling the path from the root node to $n$. If
$\substform{\alpha(n)}$ is satisfiable, the sequence $\alpha(n)$ is
feasible, in which case a model of $\substform{\alpha(n)}$ is obtained
and a word $w \in L(\A)$ is returned. Otherwise, $\alpha(n)$ is an
infeasible input sequence and the procedure enters the refinement
phase (lines \ref{ln:refine-begin}-\ref{ln:refine-end}). The GLI for
$\alpha(n)$ is used to strenghten the labels of all the ancestors of
$n$, by conjoining the formulae of the interpolant, changed according
to Lemma \ref{lemma:refinement}, to the existing labels.

In this process, the nodes on the path between $\rootNode$ and $n$,
including $n$, might become eligible for coverage, therefore we
attempt to close each ancestor of $n$ that is impacted by the
refinement (line \ref{ln:refine-end}). Observe that, in this case the
call to $\Call{Close}{}$ must uncover each node which is covered by a
successor of $n$ (line \ref{ln:close-uncover} of the $\Call{Close}{}$
function). This is required because, due to the over-approximation of
the sets of reachable configurations, the covering relation is not
transitive, as explained in \cite{McMillan06}. If $\Call{Close}{}$
adds a covering edge $(n_i,m)$ to $\lhd$, it does not have to be
called for the successors of $n_i$ on this path, which is handled via
the boolean flag $b$. Finally, if $n$ is still uncovered (it has not
been previously covered during the refinement phase) we expand $n$
(lines \ref{ln:expand-begin}-\ref{ln:expand-end}) by creating a new
node for each successor $s$ via the input event $a \in \Sigma$ and
inserting it into the worklist.

\section{Proofs of the Results in the Paper}
\label{app:proofs}

### 2019-02-07 quantifiers/body.tex
ok

The test cases shown in Table \ref{tab:experiments}, come from several
sources, namely predicate automata models (*.pa)
\cite{Farzan15,Farzan16} available online \cite{pa}, timed automata
inclusion problems ({\tt abp.ada}, {\tt train.ada}, {\tt
  rr-crossing.foada}), array logic entailments ({\tt
  array\_rota-} {\tt tion.ada}, {\tt array\_simple.ada}, {\tt
  array\_shift.ada}) and hardware circuit verification ({\tt hw1.ada},
{\tt hw2.ada}), initially considered in \cite{IosifRV16}. The {\tt
  train-simpleN.} {\tt foada} and {\tt fischer-mutexN.foada} examples
are parametric verification problems in which one checks inclusions of
the form $\bigcap_{i=1}^N\lang{A_i} \subseteq \lang{B}$, where $A_i$
is the $i$-th copy of the same template automaton.

The advantage of using FOADA over the INCLUDER \cite{includer} tool
from \cite{IosifRV16} is the possibility of having infinite alphabet
automata with hidden (local) variables, whose values are not visible
in the input. In particular, this is essential for checking inclusion
of timed automata that use internal clocks to control the computation.

>>>>

The test cases shown in Table \ref{tab:experiments}, come from several
sources, namely predicate automata models (*.pa)
\cite{Farzan15,Farzan16} available online \cite{pa}, timed automata
inclusion problems ({\tt abp.ada}, {\tt train.ada}, {\tt
  rr-crossing.foada}), array logic entailments ({\tt array\_rota-}
{\tt tion.ada}, {\tt array\_simple.ada}, {\tt array\_shift.ada}) and
hardware circuit verification ({\tt hw1.ada}, {\tt hw2.ada}),
initially considered in \cite{IosifRV16}, with the restriction that
local variables are made visible in the input. The {\tt
  train-simpleN.} {\tt foada} and {\tt fischer-mutexN.} {\tt foada}
examples are parametric verification problems in which one checks
inclusions of the form $\bigcap_{i=1}^N\lang{A_i} \subseteq \lang{B}$,
where $A_i$ is the $i$-th copy of the same template automaton.

The advantage of using FOADA over the INCLUDER \cite{includer} tool
from \cite{IosifRV16} is the possibility of having automata over
infinite alphabets with local variables, whose values are not visible
in the input. In particular, this is essential for checking inclusion
of timed automata that use internal clocks to control the computation.
